---
title: "STAT 1341 FINAL PROJECT"
author: "Olivia Snyder"
date: "2025-12-04"
output: html_document
---

# Loading and Cleaning

```{r}
library(rvest)
library(dplyr)
library(stringr)
library(ggplot2)
stats <- read.csv("stats.csv")
stats <- stats %>% 
  mutate(
    Team = str_remove(Team, "\\*"),
    Team = str_replace_all(Team, " ", "_"),
    Team = str_trim(Team)
  ) %>% 
  filter(Team != "League_Average")
nhl_url <- "https://www.hockey-reference.com/leagues/NHL_2003.html"
nhl_page <- read_html(nhl_url)
nhl_tables <- nhl_page %>% 
  html_table()
```

# Filtering + Combining

```{r}
nhl_stats <- bind_rows(nhl_tables[1], nhl_tables[2])
nhl_stats <- nhl_stats %>% 
  rename(Team = "...1",
         PTS. = "PTS%") %>% 
  select(-"RPt%") %>% 
  mutate(
    Team = str_remove(Team, "\\*"),
    Team = str_replace_all(Team, " ", "_"),
    Team = str_trim(Team)
  ) %>% 
  filter(!str_detect(Team, "Division"))

nhl <- nhl_stats %>% 
  left_join(stats, by = "Team", suffix = c("", ".nhl_dup")) %>% 
  select(-ends_with(".nhl_dup"))
nhl$Team <- gsub("Atlanta_Thrashers", "Winnipeg_Jets", nhl$Team)
nhl
```

# Adding New Statistics to Dataset

```{r}
nhl <- nhl %>%
  mutate(
    GF = as.numeric(GF),
    GA = as.numeric(GA),
    PTS. = as.numeric(PTS.),
    GP = as.numeric(GP),
    SRS =as.numeric(SRS)) %>% 
  mutate(
    PDO = ((S./100) + SV.)*100,
    STR = PP. + PK.,
    pythagorean_pct = (GF^2.15) / (GF^2.15 + GA^2.15),
    residual = PTS. - pythagorean_pct,
    goal_diff = GF - GA,
    MOV = goal_diff / GP
  )
```

# PDO

```{r}
nhl %>% 
  ggplot(aes(x=reorder(Team, PDO), y = PDO)) +
  geom_col(aes(fill = factor(PDO > 100)), show.legend = TRUE) + 
  coord_flip(ylim = c(95,105)) + labs(
    title = "Team PDO Scores",
    x = "Team",
    y = "PDO Score",
    fill = "PDO > 100 ?"
  )
```

# Special Teams Rating (STR)

```{r}
nhl %>% 
  arrange(STR) %>% 
  ggplot(aes(x= reorder(Team, STR), y = STR))+
  geom_col(aes(fill = factor(STR > 100)), show.legend = FALSE) + 
  coord_flip(ylim = c(95,110)) + labs( 
    title = "Special Teams Rating (PP% + PK%)",
    x = "Team",
    y = "STR")
```

# Simple Rating System (SRS)

```{r}
nhl %>% 
  ggplot(aes(x = SRS))+
  geom_histogram(binwidth = 0.1, fill = "purple", color = "black", alpha = 0.7) + 
  labs(
    title = "Distribution of Simp[e Rating System Scores",
    x = "SRS Scores",
    y = "Number of Teams"
  )
```

# Scoring Distributions 

```{r}
nhl %>% 
  ggplot(aes(x = GF, y = GA, label = Team)) + 
  geom_point(aes(color = goal_diff),alpha = 0.6, size = 2) +
  geom_text(nudge_y = 2.5)+
  labs(
    title = "Goals For vs. Goals Against",
    x = "Goals For",
    y= "Goals Against"
  )
nhl %>% 
  ggplot(aes(x = MOV)) +
  geom_histogram(binwidth = 0.1, fill = "orange", color = "black", alpha = 0.6) +
  labs(
    title = "Distribution of Average Margin of Victory",
    x = "Average Score Differential per Game",
    y = "Number of Teams"
  )
```

# Pythagorean Win Percentages + Residuals

```{r}
nhl %>% 
  ggplot(aes(x = reorder(Team, pythagorean_pct), y = pythagorean_pct)) +
  geom_col(fill = "pink3")+
  geom_hline(yintercept = 0.500) +
  coord_flip(ylim = c(0.35,0.7)) + 
  labs(title = "Pythagorean Win % by Team",
       x = "Team",
       y = "Pythagorean Win Percentage") 
nhl %>%
  ggplot(aes(x = as.numeric(pythagorean_pct), y = as.numeric(PTS.), label = Team)) +
  geom_point(aes(color = residual),alpha = 0.6, size = 3) +
  scale_color_gradient(
    low = "blue",
    high = "red",
    name = "Residuals"
  ) +
  geom_abline(slope = 1, intercept = 0)+
  coord_flip() +
  labs(
    title = "Pythagorean Win % vs. Actual Win %",
    x = "Pythagorean Win %",
    y= "Actual Win %"
  )
  
```

# Summary Statistics

```{r}
summary_stats <- nhl %>% 
  summarise(
    PDO_mean = mean(PDO),
    PDO_SD = sd(PDO),
    PDO_min = min(PDO),
    PDO_max = max(PDO),
    STR_mean = mean(STR),
    STR_SD = sd(STR),
    STR_min = min(STR),
    STR_max = max(STR),
    SRS_mean = mean(SRS),
    SRS_SD = sd(SRS),
    SRS_min = min(SRS),
    SRS_max = max(SRS),
    MOV_mean = mean(MOV),
    MOV_SD = sd(MOV),
    MOV_min = min(MOV),
    MOV_max = max(MOV),
    Pythagorean_mean = mean(pythagorean_pct),
    Pythagorean_SD = sd(pythagorean_pct),
    Pyth_min = min(pythagorean_pct),
    Pyth_max = max(pythagorean_pct),
    Residual_mean = mean(residual),
    Residual_SD = sd(residual),
    Residual_min = min(residual),
    Residual_max = max(residual)
  )
head(nhl %>% arrange(desc(SRS)), 1)
head(nhl %>% arrange(SRS), 1) 
head(nhl %>% arrange(desc(pythagorean_pct)), 1)
head(nhl %>% arrange(pythagorean_pct), 1)
head(nhl %>% arrange(desc(residual)), 1)
head(nhl %>% arrange(residual), 1)
summary_stats

pyth_corr <- cor(nhl$pythagorean_pct, nhl$PTS.)
pyth_corr

score_diff_cor <- cor(nhl$GF, nhl$GA)
score_diff_cor
```

# Poisson Modeling

```{r}
scores <- read.csv("nhl_data.csv")
scores <- scores %>% 
  filter(season == 2002, game_type == "r") %>% 
  select(home_team, away_team, home_score, away_score, tie, location)

scores$home_team = gsub(" ", "_", scores$home_team)
scores$away_team = gsub(" ", "_", scores$away_team)

for (i in 1:nrow(scores)){
  if(scores$tie[i]==1){
    if(scores$home_score[i] > scores$away_score[i]){
      scores$home_score[i] = scores$home_score[i] - 1}
    else if (scores$home_score[i] < scores$away_score[i]){
      scores$away_score[i] = scores$away_score[i] - 1}
  }
}

teams = sort(unique(scores$home_team))
off_teams = teams
def_teams = off_teams

data_matrix = as.data.frame(matrix(nrow = 2*nrow(scores), ncol = 2+2*length(teams)))

for (i in 1:length(off_teams)){
  off_teams[i] = paste(off_teams[i], "Off.")
}
for (i in 1:length(off_teams)){
  def_teams[i] = paste(def_teams[i], "Def.")
}

colnames(data_matrix) = c("Score", "Location", off_teams, def_teams)
data_matrix[is.na(data_matrix)] = 0

for (i in 1:nrow(scores)){
  for (j in 1:length(teams)) {
    if (scores$home_team[i] == teams[j]){
      data_matrix[i,j+2] = 1
      data_matrix[i+nrow(scores), j+2+length(teams)] = -1
    break
    }
  }

for (j in 1:length(teams)) {
  if(scores$away_team[i] == teams[j]){
    data_matrix[i+nrow(scores), j+2] = 1
    data_matrix[i, j+2+length(teams)] = -1
    break
    }
  }
}
data_matrix$Score = c(scores$home_score, scores$away_score)
data_matrix$Location = c(scores$location, 0*scores$location)

model = glm(Score~., data= data_matrix, family = "poisson")
coefficients = coef(model, complete = TRUE)
summary(model)
```

# Expected Goals and Probabilities

```{r}
max_goals = 8
ht = "New_Jersey_Devils"
at = "Anaheim_Ducks"

exp_score = matrix(nrow = max_goals+1, ncol = max_goals+1, byrow = TRUE)

for( i in 1:length(teams)){
  if(teams[i]==ht){
    ht_index = i
  }
  if(teams[i] == at){
    at_index = i
  }
}

home_score = exp(coefficients[1] + coefficients[2] + coefficients[2 + ht_index] - coefficients[2+length(teams) + at_index])
away_score = exp(coefficients[1] + coefficients[2 + at_index]- coefficients[2 + length(teams) + ht_index])

home_win_prob = 0
draw_prob = 0
away_win_prob = 0

for( i in 0:max_goals){
  for (j in 0:max_goals) {
    outcome_prob = ((home_score^i)*(exp(-home_score))/factorial(i)) * ((away_score^j)* (exp(-away_score))/factorial(j))
    exp_score[i+1, j+1] = outcome_prob
    if(i == j){
      draw_prob = draw_prob + outcome_prob
    }
    else if (i < j){
      away_win_prob = away_win_prob + outcome_prob
    }
    else{
      home_win_prob = home_win_prob + outcome_prob
    }
  }
}
scoreline_dist = round(exp_score*100, 2)
colnames(scoreline_dist) = c(0:max_goals)
rownames(scoreline_dist) = c(0:max_goals)

print(paste("Expected Home Goals (Devils):", round(home_score, 3)))
print(paste("Expected Away Goals (Ducks):", round(away_score, 3)))
print("Scoreline Probability Distribution (in %):")
print(scoreline_dist)
print(paste("Home Win Probability:", round(home_win_prob * 100, 2), "%"))
print(paste("Draw Probability:", round(draw_prob * 100, 2), "%"))
print(paste("Away Win Probability:", round(away_win_prob * 100, 2), "%"))
```

# Elo Ratings 
## HFA

```{r}
team_info <- read.csv("nhl_initial_elos.csv")
scores <- read.csv("nhl_data.csv")

scores$home_team <- gsub(" ", "_", scores$home_team)
scores$away_team <- gsub(" ", "_", scores$away_team)
team_info$team <- gsub(" ", "_", team_info$team)

non_ties = subset(scores, season == 2002 & game_type == "r" & location == 1 & home_score != away_score)

home_wins = sum(non_ties$home_score > non_ties$away_score)
games = nrow(non_ties)

home_win_prob = home_wins / games
hfa = -400 * log10(1 / home_win_prob - 1)
print(hfa)
```

## Optimal K

```{r}

starting_weight = 5
iterations = 100
step_size = 0.1
first_season = 2002

first_game_index = which(scores$season == first_season & scores$game_type == "r")[1]
last_game_index = which(scores$season == first_season & scores$game_type == "r") 
last_game_index = last_game_index[length(last_game_index)]

weights = starting_weight + (1:iterations) * step_size
errors = numeric(iterations)

for (counter in 1:iterations) {
    weight = weights[counter]
    current_team_info = team_info
    error = 0
    for (i in first_game_index:last_game_index){
       game = scores[i, ]
        home_team = game$home_team
        away_team = game$away_team
        home_score = game$home_score
        away_score = game$away_score
        location = game$location
        
        home_index = match(home_team, current_team_info$team)
        away_index = match(away_team, current_team_info$team)
        home_elo = current_team_info$rating[home_index]
        away_elo = current_team_info$rating[away_index]
        
        win_prob = 1 / (10^((away_elo - (home_elo + hfa * location)) / 400) + 1)

        if(home_score > away_score) { 
            home_result = 1  
        } else if(home_score < away_score) { 
            home_result = 0  
        } else { 
            home_result = 0.5  
        }
        error = error + (home_result - win_prob)^2
        score_diff = max(abs(game$home_score - game$away_score), 0.50)
        home_elo_adjustment = weight * sqrt(score_diff) * (home_result - win_prob)
        current_team_info$rating[home_index] = home_elo + home_elo_adjustment
        current_team_info$rating[away_index] = away_elo - home_elo_adjustment
    }

    errors[counter] = error
    print(error)
}
optimal_weight = weights[which.min(errors)]
cat("Optimal weight:", optimal_weight, "\n")

plot(weights, errors, pch = 19, xlab = "Weight", ylab = "Error")
```

## Elo Calculation

```{r} 
final_k = optimal_weight
team_to_track = "Anaheim_Ducks"
team_results = data.frame()
initial_elos_df <- team_info %>% 
    select(team, Preseason_Elo = rating)

final_elos_df = team_info
for (i in first_game_index:last_game_index) {
    game = scores[i, ]
    home_team = game$home_team
    away_team = game$away_team
    home_score = game$home_score
    away_score = game$away_score
    location = game$location
    
    home_index = match(home_team, final_elos_df$team)
    away_index = match(away_team, final_elos_df$team)
    home_elo = final_elos_df$rating[home_index]
    away_elo = final_elos_df$rating[away_index]
    
    win_prob = 1 / (10^((away_elo - (home_elo + hfa * location)) / 400) + 1)

    if (home_score > away_score) {
        home_result = 1
    } else if (home_score < away_score) {
        home_result = 0
    } else {
        home_result = 0.5
    }

    score_diff = max(abs(game$home_score - game$away_score), 0.50)
    home_elo_adjustment = final_k * sqrt(score_diff) * (home_result - win_prob)
    
    final_elos_df$rating[home_index] = home_elo + home_elo_adjustment
    final_elos_df$rating[away_index] = away_elo - home_elo_adjustment
    if(home_team == team_to_track || away_team == team_to_track)
    {
      team_is_home = (home_team == team_to_track)
      
      opponent = ifelse(team_is_home, away_team, home_team)
        pregame_elo = ifelse(team_is_home, home_elo, away_elo)
        opponent_elo = ifelse(team_is_home, away_elo, home_elo)
        team_score = ifelse(team_is_home, game$home_score, game$away_score)
        opp_score = ifelse(team_is_home, game$away_score, game$home_score)
        team_result = ifelse(team_is_home, home_result, 1 - home_result)
        team_prob = ifelse(team_is_home, win_prob, 1 - win_prob)
        
        elo_change = ifelse(team_is_home, home_elo_adjustment, -home_elo_adjustment)
        postgame_elo = pregame_elo + elo_change
        game_num = nrow(team_results) + 1
        
        team_results = rbind(team_results, data.frame(
            Game_Number = game_num,
            opponent = opponent,
            pregame_elo = round(pregame_elo, 2),
            opponent_elo = round(opponent_elo, 2),
            win_probability = round(team_prob, 4),
            result = team_result,
            team_score = team_score,
            opponent_score = opp_score,
            elo_adjustment = round(elo_change, 2),
            postgame_elo = round(postgame_elo, 2)
        ))
    }
}
```

## Elo Visualizations

```{r}
elo_graph_one_team = team_results %>%
  ggplot(aes(x = Game_Number, y = postgame_elo)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = paste("Elo Rating Progression for the", gsub("_", " ", team_to_track), " (2002-2003)"),
    x = "Game Number in Season",
    y = "Elo Rating"
  ) +
  theme_minimal()

print(elo_graph_one_team)

comparison_df <- final_elos_df %>% 
    select(team, Postseason_Elo = rating) %>%
    left_join(initial_elos_df, by = "team") %>%
    mutate(Elo_Change = Postseason_Elo - Preseason_Elo) %>%
    arrange(desc(Postseason_Elo)) %>% 
    filter(Elo_Change != 0.00000)

comparison_plot <- comparison_df %>%
    ggplot(aes(x = reorder(team, Postseason_Elo), y = Elo_Change)) +
    geom_col(aes(fill = Elo_Change > 0), show.legend = FALSE) +
    geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    coord_flip() +
    labs(
        title = "Change in Elo Rating from Preseason to Postseason (2002-2003)",
        x = "Team",
        y = "Elo Change)",
        subtitle = paste("Optimal K-factor (Weight):", round(optimal_weight, 2))
    ) +
    scale_fill_manual(values = c("TRUE" = "green4", "FALSE" = "firebrick")) +
    theme_minimal()

print(comparison_plot)
print(comparison_df %>% 
        arrange(team))
```

# Monte Carlo Simulation

```{r}
weight = 12 
hfa = 39.34695
divisions = list(
    Atlantic = c("New_Jersey_Devils", "Philadelphia_Flyers", "New_York_Islanders", "New_York_Rangers", "Pittsburgh_Penguins"),
    Northeast = c("Ottawa_Senators", "Toronto_Maple_Leafs", "Boston_Bruins", "Montreal_Canadiens", "Buffalo_Sabres"),
    Southeast = c("Tampa_Bay_Lightning", "Washington_Capitals", "Florida_Panthers", "Carolina_Hurricanes", "Columbus_Blue_Jackets"),
    Central = c("Detroit_Red_Wings", "St._Louis_Blues", "Chicago_Blackhawks", "Nashville_Predators", "Minnesota_Wild"),
    Northwest = c("Colorado_Avalanche", "Vancouver_Canucks", "Edmonton_Oilers", "Calgary_Flames", "Winnipeg_Jets"),
    Pacific = c("Dallas_Stars", "Anaheim_Ducks", "Los_Angeles_Kings", "Utah_Hockey_Club", "San_Jose_Sharks")
)

team_division_map = data.frame(
  team = unlist(divisions),
  division = rep(names(divisions), sapply(divisions, length))
)
team_info = team_info %>% 
    select(team, rating) %>% 
    left_join(team_division_map, by = "team")

season_schedule = scores[first_game_index:last_game_index, ]

set.seed(31)
iterations = 1000
results = data.frame(matrix(0, ncol = 6, nrow = nrow(team_info)))
colnames(results) = c("team", "starting_elo", "ending_elo", "actual_wins", "average_wins", "division_titles")
results$team = team_info$team
results$starting_elo = team_info$rating
simulated_win_totals = data.frame(matrix(0, ncol = nrow(team_info), nrow = iterations))
colnames(simulated_win_totals) = team_info$team
for(i in 1:iterations) {
    if(i %% 1000 == 0) {print(i)}
    season_stats = team_info %>% mutate(wins = 0, rand = runif(n()))
    
    for(j in 1:nrow(season_schedule)) {
        game = season_schedule[j, ]
        home_team = game$home_team
        away_team = game$away_team
        home_score = game$home_score
        away_score = game$away_score
        location = game$location
        
        home_index = match(home_team, season_stats$team)
        away_index = match(away_team, season_stats$team)
        home_elo = season_stats$rating[home_index]
        away_elo = season_stats$rating[away_index]
        home_win_prob = 1 / (1 + 10^((away_elo - (home_elo + hfa * location)) / 400))
        u = runif(1)
        if(u < home_win_prob) {
            season_stats$wins[home_index] = season_stats$wins[home_index] + 1
        } else {
            season_stats$wins[away_index] = season_stats$wins[away_index] + 1
        }
        if(home_score > away_score) {
            home_result = 1
        } else if(home_score < away_score) {
            home_result = 0
        } else {
            home_result = 0.5
        }
        score_diff = max(abs(home_score - away_score), 0.5)
        home_elo_adjustment = weight * sqrt(score_diff) * (home_result - home_win_prob)
        season_stats$rating[home_index] = home_elo + home_elo_adjustment
        season_stats$rating[away_index] = away_elo - home_elo_adjustment
        
        if(i == 1) {
            if(home_score > away_score) {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 1
            } else if(home_score < away_score) {
                results$actual_wins[away_index] = results$actual_wins[away_index] + 1
            } else {
                results$actual_wins[home_index] = results$actual_wins[home_index] + 0.5
                results$actual_wins[away_index] = results$actual_wins[away_index] + 0.5
            }
        }
    }
    results$average_wins = results$average_wins + season_stats$wins
    simulated_win_totals[i,] = t(season_stats$wins)
    division_winners = data.frame(matrix(ncol = 6, nrow = 0))
    colnames(division_winners) = colnames(season_stats)
    for(div in names(divisions)) {
        div_standings = season_stats[which(season_stats$division == div),]
        div_standings = div_standings[order(-div_standings$wins, -div_standings$rand),]
        division_winners = rbind(division_winners, div_standings[1,])
    }
    div_winner_indices = match(division_winners$team, results$team)
    results$division_titles[div_winner_indices] = results$division_titles[div_winner_indices] + 1
}
results$ending_elo = season_stats$rating
results$average_wins = results$average_wins/iterations
results$division_titles = results$division_titles/iterations

print(results %>% arrange(desc(average_wins)))
```

## Visualizaion

```{r}

team_to_plot = "Carolina_Hurricanes"
devils_sim_wins = simulated_win_totals[[team_to_plot]]

hist(devils_sim_wins,
     main = paste("Simulated Win Totals for", gsub("_", " ", team_to_plot)),
     xlab = "Simulated Wins",
     ylab = "Frequency",
     col = "coral",
     border = "black")

results %>% 
  filter(average_wins != 0) %>% 
  ggplot(aes(x = average_wins, y = actual_wins, label = team))+
  geom_point(alpha = 0.6, size = 3) +
  geom_abline(slope = 1, intercept = 0, col = "red") + 
  geom_text(nudge_y = 1) +
  labs(
  title = "Simulated Wins vs. Actual Wins",
  x = "Simulated Wins",
  y = "Actual Wins"
)
cor(results$actual_wins, results$average_wins)
results$residual <- results$actual_wins - results$average_wins
```

